%&latex
\section{External Dependency Management}

A programming language's ability to aid the programmer in managing external
dependencies---by, for instance, providing various tools that come installed
with its compiler or development environment---is generally not incorporated
into that language's syntax or, by extension, its semantics. As a result, it is
traditionally not considered an aspect of that language's feature set per se.
Notwithstanding, external dependency management is more and more becoming an
appreciated addition to the \textit{ecosystem} around a language, especially
so for compiled languages. In fact, many would consider automated external
dependency management to be a must for any modern programming language. As
it directly affects both the portability of code, as well as the (ease of)
managing different versions of a dependency, a language's ability to unburden
the programmer from the manual management of external dependencies can greatly
improve the maintainability of a project, and can therefore indeed be viewed as
a feature of the language in light of its \textit{pragmatics}. In this section,
we will be discussing the practicalities of working with external code in both
Rust and \cpp. We will do this by means of a comparison between two window
manager implementations, one written in Rust, and the other written in \cpp,
which we will henceforth be referring to as \wmrs and \wmcpp, respectively.

Both window managers are built on top of the X Window System, which means
they rely on an external library to interface with the X server\cite{x11}.
Although \wmrs and \wmcpp each use a different library to achieve this (XCB over
\ttt{libxcb}\cite{xcb} and Xlib over \ttt{libX11}\cite{xlib}, respectively), the
methodology is the same, as they both require the importing of an external body
of code.

\cppsubsection{Ad Hoc Dependency Management}

\cpp does not come with an external dependency manager. That is to say, the
official ISO standard \cpp specification\cite{cppstd} does not outline the
functionality or otherwise existence of any \textit{package manager}. As a
result, developers have no other choice than to resort to third-party tools
to manage dependencies, to reinvent the wheel themselves and hack together
configure and build scripts that take external dependencies into account, or to
disregard package management altogether and let other developers and end users
resolve dependencies on their own. In any case, \cpp dependency management is
complex, and the lack of a standardized tool inhibits the adoption of \cpp
projects, as well as their portability to other platforms and development
environments.

Since the only external dependency \wmcpp is reliant on is \ttt{libX11}, which
is fairly ubiquitous and readily available through distributions' own package
managers, it merely links with the X11 development libraries without the use of
any form of own dependency management. Specifically, building the project is
done using the following \textup{Make} script (large parts altered or redacted
for clarity):

\begin{mintedblock}{make}
  CXXFLAGS = -std=c++17 -march=native -O3
  LDFLAGS = `pkg-config --libs x11` -flto

  obj/%.o: src/%.cc
    g++ ${CXXFLAGS} -MMD -c $< -o $@
  release: obj/%.o
    g++ $< ${LDFLAGS} -o bin/wmCPP
\end{mintedblock}

Assuring the availability of an external body of code is not the only aspect
of external dependecy management that affects the pragmatics of \cpp. Writing
a library that is to be imported as an external dependency and including code
from an external dependency both require special care to be taken to make sure
no symbol collisions occur. In header files, this is usually done by making
use of so-called \textit{header guards}, a set of preprocessor directives that
render including a header an idempotent operation, preventing double inclusion
errors. If incorrect double inclusion does unexpectedly occur, for instance
due to collisions in the header guards themselves, vague---as header guards
are handled by the preprocessor, not the compiler---and hard-to-trace compiler
errors arise, subjecting the programmer to unnecessary mental strain, and
wasting time. Although not part of the official standard\cite{cppstd}, many \cpp
compilers support the \cppin{#pragma once} preprocessor directive, providing the
same functionality as header guards, but preventing the occurrence of double
inclusion errors.

Another source of issues is the order of include directives in \cpp. The
ordering of included files relative to one another can significantly influence
the outcome of the compilation stage, possibly causing errors, one of which
being the introduction of circular dependencies while there actually aren't any
there (e.g. a forward declaration that was included only after the circular
dependency was already introduced).

% https://doc.rust-lang.org/cargo/guide/

\rssubsection{Automated Dependency Management using Cargo}

The Rust programming language has its own package manager, called
\textup{Cargo}. \textup{Cargo} can automatically download a project's external
dependencies (and, transatively, \textit{their} dependencies), compile them, and
install them locally, such that they can be used during the linking stage of the
build process\cite{thecargobook}. To make proper use of \textup{Cargo} within
a Rust project, that project must be turned into a so-called \textit{package}.
Rust packages are simply a collection of source files along with a manifest file
(named \ttt{Cargo.toml}) in the project's root directory. This manifest file
describes the package's meta-information (such as its name and version), and a
set of \textit{target crates}\cite{thecargobook}. A crate is the source code
or compiled artifact of either a library or executable program, or possibly a
compressed package that is grabbed from a registry (a service that provides a
collection of downloadable crates)\cite{thecargobook}. A package's manifest file
describes each of its target crates by specifying their type (binary executable
or library), their metadata, and how to build them\cite{thecargobook}.

The following manifest file describes the package that represents our Rust
window manager, \wmrs (parts redacted for clarity):

\begin{mintedblock}{shell}
  [package]
  name = "wmRS"
  edition = "2018"
  description = """
  An ICCCM & EWMH compliant X11 reparenting,
  tiling window manager, written in Rust
  """
  [profile.release]
  lto = true
  [lib]
  name = "winsys"
  path = "src/winsys/mod.rs"
  [[bin]]
  name = "wmRS"
  path = "src/core/main.rs"
  [[bin]]
  name = "wmRSbar"
  path = "src/bar/main.rs"
  [[bin]]
  name = "wmRSclient"
  path = "src/client/main.rs"
  [dependencies]
  x11rb = {
    version = "0.8.0",
    features = [
      # ...
    ]
  }
\end{mintedblock}

% https://doc.rust-lang.org/book/ch10-02-traits.html
% https://doc.rust-lang.org/rust-by-example/trait.html
% https://blog.logrocket.com/rust-traits-a-deep-dive/
% https://doc.rust-lang.org/book/ch19-03-advanced-traits.html

The package consists of a single library crate (\ttt{winsys}), along with three
binary executable crates (\ttt{wmRS}, \ttt{wmRSbar}, and \ttt{wmRSclient}).

\rssubsection{Decoupling External Dependencies using Traits}

\textsc{wmRS}'s \rsin{winsys} library is an abstraction above and wrapper
around the API into the underlying windowing system. It defines a single Rust
\textit{trait} that represents the connection between the window manager and
some windowing system. A trait is a \textit{zero-overhead}\cite{rustblogtraits}
collection of methods that is declared for some (at declare-time) unknown type
\rsin{Self} (which at implementation-time becomes the implementing type), most
often used with the intention to implement shared behavior\cite{therustbook,
rustbyexample}. A comparable concept from other languages that most closely
resembles traits are \textit{interfaces}, although there are differences
still\cite{therustbook, rusttraitsdeepdive}. A small snippet from \wmrs's
\ttt{winsys} library \rsin{Connection} trait looks as follows.

% TRAITS VS CONCEPTS
% https://www.modernescpp.com/index.php/c-20-concepts-the-details
% https://www.quora.com/What-are-the-similarities-and-differences-between-C-s-concepts-and-Rusts-traits

\begin{rustblock}
  pub trait Connection {
    fn step(&self) -> Option<Event>;
    fn move_window(&self, window: Window, pos: Pos);
    fn resize_window(&self, window: Window, dim: Dim);
    fn close_window(&self, window: Window);
    // ...
  }
\end{rustblock}

Here, we've supplied a set of function prototypes that eventually become
the methods that every type that implements this trait will be required to
define. Although we haven't done so here, traits themselves are also allowed
to define their functions with some default behavior, that may or may not
be overwritten by its implementors\cite{therustbook, rusttraitsdeepdive}.
Our \rsin{Connection} trait represents shared behavior that every windowing
system wrapper is required to contain. This allows for the decoupling of the
implementation of the higher-level window management functionality from that of
the interface with the windowing system (e.g. the explicit drawing of primitives
to the screen, or the management and manipulation of windowing system specific
window representations). It additionally allows for the seamless transition from
one windowing system to another, as multiple windowing systems can be targeted
by implementing the trait, effectively creating a new wrapper around an external
library that directly interfaces that windowing system. Currently, \wmrs only
implements the interface with the X Window System, but interfacing with the
newer and more modern Wayland\cite{wayland} is as easy as implementing a new
\rsin{Connection} to it. \wmrs's implementation for this trait, targeting the X
Window System, is partly given as follows:

\begin{rustblock}
  use x11rb::connection;
\end{rustblock}
\begin{rustblock}
  pub struct XConnection
    <'conn, Conn: connection::Connection>
  {
    conn: &'conn Conn,
    // ...
  }
\end{rustblock}
\begin{rustblock}
  impl<'conn, Conn: connection::Connection>
    Connection for XConnection<'conn, Conn>
  {
    #[inline]
    fn step(&self) -> Option<Event> { /* ... */ }
    // ...
  }
\end{rustblock}

The \rsin{XConnection} structure introduces two generic type arguments, one
being a lifetype parameter (\rsin{'conn}), and the other being the type
of the struct's \rsin{conn} field (\rsin{Conn}), which \textit{requires}
the \rsin{connection::Connection} trait to be implemented. This
\rsin{connection::Connection} trait is imported from the external package
\ttt{x11rb}, which we supplied as a project dependency in our manifest file. The
\ttt{x11rb} package serves as a Rust API to the X Window System, essentially
providing us with Rust bindings to interact with the X server. Finally, the
\rsin{XConnection} struct implements our previously defined \rsin{Connection}
trait, allowing it to be used by our window manager by means of composition.
The listing below shows the structure that represents and encapsulates
\textsc{wmRS}'s core window manager logic. It \textit{contains} a reference to
\textit{some} type that implements our \rsin{Connection} trait (i.e. a wrapper
around \textit{some} windowing system).

\begin{rustblock}
  pub struct Model<'model> {
    conn: &'model mut dyn Connection,
    // ...
  }
\end{rustblock}

% https://doc.rust-lang.org/1.8.0/book/trait-objects.html#static-dispatch
% https://doc.rust-lang.org/1.8.0/book/trait-objects.html#dynamic-dispatch
% https://doc.rust-lang.org/book/ch10-01-syntax.html#performance-of-code-using-generics

Since we're using polymorphism here to abstract away from the actual,
concrete implementation of the connection type, Rust needs to be able to
determine at runtime which specific version it should actually run (known as
\textit{dispatch})\cite{therustbook, rustblogtraits}. It can do so either
\textit{statically} or \textit{dynamically}. Static dispatch in Rust makes
use of \textit{monomorphization} at compile time to convert generic code into
``specific'' code: one version for each concrete type that is used as a generics
argument\cite{therustbook, rustblogtraits}. That means that at runtime, no time
overhead is incurred when running generic code, as the specific versions of the
code to run are baked into the binary\cite{therustbook, rustblogtraits}. An
example of static dispatch in \wmrs is the following \rsin{Cycle} struct, which
allows for cycling forward and backward through a collection of generic items.

\begin{rustblock}
  pub struct Cycle<T>
  where
    T: Identify + Debug,
  { /* ... */ }
\end{rustblock}
\begin{rustblock}
  impl<T> Cycle<T>
  where
      T: Identify + Debug,
  { /* ... */ }
\end{rustblock}

This struct is used to cycle through both the workspace structures managed
by the window manager, as well as the clients within a workspace. That
means two specific versions of this struct are constructed at compile-time:
one for \rsin{Cycle<Workspace>}, and one for \rsin{Cycle<Client>}, where
\rsin{Workspace} and \rsin{Client} in turn both implement the \rsin{Identify}
and \rsin{Debug} traits (as per the constraint on \rsin{T}).

% https://doc.rust-lang.org/book/ch19-04-advanced-types.html#dynamically-sized-types-and-the-sized-trait
% https://doc.rust-lang.org/book/ch17-02-trait-objects.html

Dynamic dispatch defers resolving generic code until it is required at
runtime, making use of so-called \textit{trait objects}\cite{therustbook,
rustdynamicdispatch}. A trait object is an opaque value of a type that
implements some set of traits\cite{therustreference}. It's opaque,
as one cannot know which concrete type is behind a trait object's
pointer up front\cite{therustreference}. Whereas the size of each
monomorphized type is always known, trait objects are therefore dynamically
sized\cite{therustreference}. To resolve a call to one of such an opaque
type's methods at runtime, \textit{virtual method tables} (\textit{vtables})
are used\cite{therustreference}. Each instance of a pointer to a trait object
consists of a pointer to an instance of some type \rsin{T} that implements
the set of traits, as well as a pointer to a vtable that contains a function
pointer to \rsin{T}'s implementation of each method of the implemented set of
traits (and their supertraits)\cite{therustreference}. Our \rsin{Model} struct's
\rsin{conn} field is a pointer to such a trait object. Behind the pointer, we
can have any implementation of the \rsin{Connection} trait. Which of the trait's
implementors' methods are called, is determined at runtime.

Looking back at \wmrs's manifest file, the three binary executable crates in
its package respectively represent the window manager itself, a client program
to communicate with the window manager (to control various window management
activities, such as closing the currently focused window, from scripts or the
command line), and a status bar that displays information about the state of the
window manager, such as the currently activated workspace. Each make use of the
\ttt{winsys} library to communicate with the underlying windowing system.

\cppsubsection{Abstraction through Abstract Classes}

Our \cpp window manager implementation, \wmcpp, attempts to achieve the same
flexibility by making use of \textit{abstract classes}. Abstract classes define
abstract types that cannot themselves be implemented, but are instead used
to establish a common denominator between types that should present shared
behavior. Abstract classes can mimick the behavior of interface constructs in
languages such as Java by declaring only \textit{pure virtual} methods. Pure
virtual methods are methods that do not expose any associated inline logic, and
as such \textit{must} be implemented by any inheriting subclasses. Consider
\wmcpp's \cppin{Connection} abstract class.

% https://en.cppreference.com/w/cpp/language/abstract_class

\begin{cppblock}
  class Connection
  {
  public:
    virtual ~Connection() {}
\end{cppblock}
\begin{cppblock}
    virtual Event step() = 0;
    virtual void move_window(Window, Pos) = 0;
    virtual void resize_window(Window, Dim) = 0;
    virtual void close_window(Window) = 0;
    // ...
  };
\end{cppblock}

The fact that this class contains \textit{at least} a single virtual method
declaration without an inline implementation (i.e., its declaration appears
to be assigned to zero), makes it an abstract class. When not a single
method (except for possibly its constructor or destructor) has an inline
implementation, that class is considered to be a proper interface.

Although an abstract class's pure virtual methods cannot be called
\textit{dynamically} (i.e., using virtual dispatch), it may still implement
associated logic that can subsequently be called \textit{statically}. Consider
part of \cppin{Connection}'s implementation.

% https://stackoverflow.com/a/12855410/11069175
% https://stackoverflow.com/questions/5481941/c-pure-virtual-function-have-body

\begin{cppblock}
  #include "connection.hh"
  #include "log.hh"
\end{cppblock}
\begin{cppblock}
  // ...
  void
  Connection::focus_window(Window window)
  {
    Logger::log_info("Focusing window %s.",
      window.to_string());
    // ...
  }
  // ...
\end{cppblock}

Calling a virtual function statically (non-virtually) is done by using its
qualified name in the call. This is especially useful for implementors (derived
classes) that all share an identical portion of code. In our example above,
regardless of the underlying windowing system, the logging of an event is a
fixed procedure, and can thus be part of the abstract class's implementation.
Performing the call from \wmcpp's \cppin{XConnection} class would look as
follows.

\begin{cppblock}
  #include "connection.hh"
\end{cppblock}
\begin{cppblock}
  // ...
  class XConnection final: public Connection
  {
  public:
    // ...
    void focus_window(Window window) override {
      // non-virtual call
      Connection::focus_window(window);
      // ...
    }
    // ...
  };
  // ...
\end{cppblock}

As can be seen in the above example, providing a concrete implementation for
our abstract notion of a connection is done through \textit{inheritance}.
Each wrapper around the connection with \textit{some} windowing system will
be represented as a class that inherits (derives) from the \cppin{Connection}
abstract class, providing an implementation specific to that windowing system.

\wmcpp's \cppin{Model} class will then \textit{contain} a reference to
\textit{some} implementation of \cppin{Connection}, as follows.

\begin{cppblock}
  #include "connection.hh"
\end{cppblock}
\begin{cppblock}
  // ...
  class Model final
  {
  public:
    Model(Connection& conn): conn(conn) {
      // ...
    }
    // ...
  private:
    Connection& conn;
    // ...
  };
\end{cppblock}

Just as in our Rust implementation, \cppin{conn}'s implementation details will
be resolved only at runtime, when they are needed, through dynamic dispatch
(making use of a similar vtable mechanism).

% https://blog.rust-lang.org/2015/05/11/traits.html
% https://doc.rust-lang.org/book/ch17-01-what-is-oo.html#inheritance-as-a-type-system-and-as-code-sharing
% https://doc.rust-lang.org/rust-by-example/generics/bounds.html

\subsection{External Dependency Management: Remarks}

While in our window manager implementations both Rust's traits and \cpp's
abstract classes appear to achieve the same objective in similar fashion, the
constructs themselves are very different. For one, implementation (traits)
is not the same as inheritance (abstract classes). By design, Rust does not
allow for any type of inheritance (i.e. one object can inherit from another
object's definition). The traditional benefits to inheritance are twofold:
code reuse and polymorphism. Rust's traits attain the same functionality
by combining implementation with \textit{generics} and \textit{trait
bounds}\cite{therustbook, rustblogtraits}. Traits allow for the central
declaration (and possibly definition with their default implementations) of
common behavior\cite{therustbook, rustblogtraits}. The use of generics on top
of traits introduces the possibility for abstractions over different possible
types\cite{therustbook, rustblogtraits}. Trait bounds then \textit{constrain}
these type abstractions, imposing restrictions on what exactly those types
are to provide\cite{therustbook, rustblogtraits, therustreference}. \wmrs's
\rsin{Cycle} trait introduces a generic type, \rsin{T}, that is \textit{bound}
by the restriction \rsin{Identify + Debug}, stipulating that whatever concrete
\rsin{T} gets passed in \textit{must} implement both \rsin{Identify} and
\rsin{Debug}.

Another useful feature of traits is the ability to implement crate-owned
traits on external (and even built-in) types. Take our \rsin{Identify} trait
as an example, which declares an \rsin{id} function that serves to uniquely
identify an implementing object, and which is additionally a bound on the types
passed into the \rsin{Cycle} struct. If we want our \rsin{Cycle} to work on
the built-in 32-bit unsigned integer type, it would first need to implement
\rsin{Identify}, which can be done as follows.

\begin{rustblock}
  impl Identify for u32 {
    #[inline(always)]
    fn id(&self) -> Ident {
      *self
    }
  }
\end{rustblock}

\cpp's abstract classes, on the other hand, do not allow for (re)derivation by
already existing (external) types, without first creating some kind of wrapper
object around them. This makes working with external code a lot more convenient
in Rust compared to \cpp.

As inheritance often leads to the sharing of more code than necessary (due to
the fact that \textit{all} of a parent class's characteristics are inherited),
many would consider it to be a suboptimal solution. Rust, with its trait
objects, allows for tighter encapsulation, and altogether presents a more modern
and flexible approach in that regard.
